---
title: 'edX HarvardX: PH125.8x | MovieLens Project'
author: "Nicole Yong"
date: "8/26/2019"
output: pdf_document
---

# OVERVIEW

## INTRODUCTION
This MovieLens project is created to fulfil the coursework requirements of the capstone module in edX HarvardX: PH125.9x (Data Science Professional Certificate).

### MovieLens Dataset
The MovieLens dataset, collected by GroupLens Research, includes movie rating datasets from the MovieLens website (http://movielens.org).  

For the purpose of this project, the MovieLens 10M Dataset is used. This dataset includes 10 million ratings applied to 10,000 movies by 72,000 users. It was released in 2009.

This dataset can be retrieved in the following links:

* https://grouplens.org/datasets/movielens/10m/
* http://files.grouplens.org/datasets/movielens/ml-10m.zip

### Goal of MovieLens Project
The goal of this project is to create a machine learning algorithm that predicts movie ratings in the validation dataset as if they were unknown.  

RMSE (Root Mean Squared Error) value is used to evaluate the effectiveness of the models. RMSE is the standard deviation of the residuals (prediction errors). Residuals are a measure of how far from the regression line data points are. Hence, a lower RMSE value indicates a better fit.

### Key Steps Taken
1. Data Gathering & Loading the Dataset
* Load the dataset from data source
* Split the dataset into train and test dataset
2. Data Preprocessing
* Explore both the train and test dataset
* Check for variables that require preprocessing (e.g. missing values, wrong format)
3. Data Exploration
* Gain insights on the dataset
* Explore the relationships between different variables
4. Model Selection and Training
* Test out different models
* Improve models' performance using insights from data exploration

# METHODS & ANALYSIS OF MOVIELENS DATASET

## Load required libraries

```{r, echo=FALSE}
library(tidyverse)
library(caret)
library(data.table)
library(lubridate)
library(dplyr)
library(caret)
library(ggplot2)
library(tidyr)
library(hrbrthemes)
library(wordcloud)
library(RColorBrewer)
```

## Print session information
To help with code reproducibility, print version information about R, the OS and attached or loaded packages. 
```{r, echo = FALSE}
sessionInfo()
```

## STEP 1: DATA GATHERING & LOADING THE DATASET

### Load MovieLens dataset
```{r}
dl <- tempfile()
download.file("http://files.grouplens.org/datasets/movielens/ml-10m.zip", dl)

ratings <- fread(text = gsub("::", "\t", readLines(unzip(dl, "ml-10M100K/ratings.dat"))),
                 col.names = c("userId", "movieId", "rating", "timestamp"))

movies <- str_split_fixed(readLines(unzip(dl, "ml-10M100K/movies.dat")), "\\::", 3)
colnames(movies) <- c("movieId", "title", "genres")
movies <- as.data.frame(movies) %>% mutate(movieId = as.numeric(levels(movieId))[movieId],
                                           title = as.character(title),
                                           genres = as.character(genres))

movielens <- left_join(ratings, movies, by = "movieId")
```

### Create train (edx) and test (validation) sets

Validation set will be 10% of MovieLens dataset

```{r}
set.seed(1)
test_index <- createDataPartition(y = movielens$rating, times = 1, p = 0.1, list = FALSE)
edx <- movielens[-test_index,]
temp <- movielens[test_index,]
```

Ensure userId and movieId in validation set are also in edx set

```{r}
validation <- temp %>% 
  semi_join(edx, by = "movieId") %>%
  semi_join(edx, by = "userId")
```

Add rows removed from validation set back into edx set

```{r, message = FALSE}
removed <- anti_join(temp, validation)
edx <- rbind(edx, removed)
rm(dl, ratings, movies, test_index, temp, movielens, removed)
```

## STEP 2: DATA PREPROCESSING

### Take a look at the train (edx) dataset

```{r}
glimpse(edx)
summary(edx)
head(edx)
```

#### Insights
1. No missing variables from summary table

2. Timestamp variable needs to be processed (converted to year)
```{r}
edx <- mutate(edx, timestamp_year = year(as_datetime(timestamp)))
```

3. Genres variable needs to be processed (separate movies with multiple genres into individual rows per genre)
```{R}
edx <- edx %>% separate_rows(genres, sep ="\\|")
```

### Take a look at the test (validation) dataset

```{r}
glimpse(validation)
summary(validation)
head(validation)
```

#### Insights
1. No missing variables from summary table

2. Timestamp variable needs to be processed (converted to year)
```{r}
validation <- mutate(validation, timestamp_year = year(as_datetime(timestamp)))
```

3. Genres variable needs to be processed (separate movies with multiple genres into individual rows per genre)
```{R}
validation <- validation %>% separate_rows(genres, sep ="\\|")
```

## STEP 3: DATA EXPLORATION

### Explore number of unique movies and users in edx training set
* There are `r n_distinct(edx$userId)` unique users
* There are `r n_distinct(edx$movieId)` unique movies

### Explore distribution of ratings in edx training set
```{r, echo=FALSE, fig.align='center', fig.width=5, fig.height=3, message=FALSE}
edx %>%
  ggplot(aes(rating)) +
  geom_histogram(binwidth = 0.5, fill="#69b3a2", color="#e9ecef") +
  scale_x_discrete(limits = c(seq(0.5,5,0.5))) +
  scale_y_continuous(breaks = c(seq(0, 10000000, 1000000))) + 
  theme(plot.title = element_text(size=15), 
        axis.title = element_text(size=10),
        axis.text.x = element_text(size=8),
        axis.text.y = element_text(size=8)) +
  ggtitle("Distribution of Ratings (Training Set)") +
  xlab("Movie Ratings") + 
  ylab("Number of Ratings")
```

### Explore distribution of ratings in validation set
```{r, echo=FALSE, fig.align='center', fig.width=5, fig.height=3, message=FALSE}
validation %>%
  ggplot(aes(rating)) +
  geom_histogram(binwidth = 0.5, fill="#69b3a2", color="#e9ecef") +
  scale_x_discrete(limits = c(seq(0.5,5,0.5))) +
  scale_y_continuous(breaks = c(seq(0, 1000000, 100000))) +
  theme(plot.title = element_text(size=15), 
        axis.title = element_text(size=10),
        axis.text.x = element_text(size=8),
        axis.text.y = element_text(size=8)) +
  ggtitle("Distribution of Ratings (Validation Set)") +
  xlab("Movie Ratings") + 
  ylab("Number of Ratings")
```

#### Insights
* Both train and test datasets have similar distribution of ratings
* More full ratings are given compared to half (0.5) ratings

### Explore number of ratings per unique movie
```{r, echo=FALSE, fig.align='center', fig.width=5, fig.height=3, message=FALSE}
edx %>%
  count(movieId) %>%
  ggplot(aes(n)) +
  geom_histogram(bins = 30, binwidth = 0.5, fill="#69b3a2", color="#e9ecef") +
  scale_x_log10() +
  theme(plot.title = element_text(size=15), 
        axis.title = element_text(size=10),
        axis.text.x = element_text(size=8),
        axis.text.y = element_text(size=8)) +
  ggtitle("Ratings per Movie") +
  xlab("Unique Movies") + 
  ylab("Number of Ratings")
```

### Explore number of ratings per unique user
```{r, echo=FALSE, fig.align='center', fig.width=5, fig.height=3, message=FALSE}
edx %>%
  count(userId) %>%
  ggplot(aes(n)) +
  geom_histogram(bins = 30, binwidth = 0.5, fill="#69b3a2", color="#e9ecef") +
  scale_x_log10() +
  theme(plot.title = element_text(size=15), 
        axis.title = element_text(size=10),
        axis.text.x = element_text(size=8),
        axis.text.y = element_text(size=8)) +
  ggtitle("Ratings per User")+
  xlab("Unique Users") + 
  ylab("Number of Ratings")
```

#### Insights
* Certain movies have a higher number of ratings compared to others
* There are some movies with very few ratings
* Certain users rate more movies than other users

### Explore the genres of movies rated
```{r, echo=FALSE, warning=FALSE, fig.align='center', fig.width=8, fig.height=3, message=FALSE}
count_genre <- edx %>% group_by(genres) %>% summarize(count_genre = n())
wordcloud(words = count_genre$genres, freq = count_genre$count_genre, min.freq = 1,
          max.words=200, random.order=FALSE, rot.per=0.35, 
          colors=brewer.pal(8, "Spectral"))
```

#### Insights
* Certain genres of movies garner more movie ratings
* However, note that a single movie can have multiple genres

## STEP 4: MODEL SELECTION AND TRAINING
Define RMSE (Root Mean Squared Error)
```{r}
RMSE <- function(true_ratings, predicted_ratings){
  sqrt(mean((true_ratings - predicted_ratings)^2))
}
```

### Part 4.1 - Baseline RMSE
To compute the baseline RMSE with the average rating computed from the training dataset
```{r}
mu <- mean(edx$rating)
baseline_RMSE <- RMSE(edx$rating, mu)
```

Test the baseline model against the validation (test) dataset  
Add baseline RMSE result to a table for later comparison
```{r, warning=FALSE, echo=FALSE}
baseline_RMSE <- RMSE(validation$rating, mu)

RMSE_results <- data_frame(Model = "Baseline RMSE", RMSE = baseline_RMSE)
RMSE_results %>% knitr::kable()
```

#### Part 4.1 Insights
* Baseline RMSE is **`r baseline_RMSE`** 
* The goal of the subsequent models is to perform better than the baseline RMSE 
* To improve subsequent models using insights gained from data exploration

#### Data Exploration Insights
* The top 3 ratings given by users are 4.0, 3.0 and 5.0 respectively 
* 0.5 (or half) ratings are less popular than whole ratings 
* Potential movie effect: Different movies have different ratings frequency and different ratings 
* Potential user effect: Certain users have rated more movies than other users

### Part 4.2 - Movie Effect
* Recap: Some movies have a higher (than mean) rating
* Account for estimated deviation of each movie's mean rating from total average mean

Account for Movie Effect
```{r}
movie_mu <- edx %>%
  group_by(movieId) %>%
  summarize(b_i = mean(rating - mu))
```

Test the adjusted model  
Add Movie Effect RMSE results to a table for later comparison
```{r, warning=FALSE, echo=FALSE}
predicted_ratings <- mu +  validation %>%
  left_join(movie_mu, by='movieId') %>%
  pull(b_i)

adj_model_1_RMSE <- RMSE(validation$rating, predicted_ratings)

RMSE_results <- bind_rows(RMSE_results,
                          data_frame(Model="Movie Effect",  
                                     RMSE = adj_model_1_RMSE))
RMSE_results %>% knitr::kable()
```

#### Part 4.2 Insights
RMSE has improved to **`r adj_model_1_RMSE`**

### Part 4.3 - User Effect
* Recap: Different users rate movies differently 
* Account for movie effect and user effect in the predicted ratings
```{r}
user_mu <- edx %>%
  left_join(movie_mu, by='movieId') %>%
  group_by(userId) %>%
  summarize(b_u = mean(rating - mu - b_i))
```

Test the adjusted model  
Add User Effect RMSE results to a table for later comparison
```{r, warning=FALSE, echo=FALSE}
predicted_ratings <- validation%>%
  left_join(movie_mu, by='movieId') %>%
  left_join(user_mu, by='userId') %>%
  mutate(pred = mu + b_i + b_u) %>%
  pull(pred)

adj_model_2_RMSE <- RMSE(validation$rating, predicted_ratings)

RMSE_results <- bind_rows(RMSE_results,
                          data_frame(Model="Movie Effect & User Effect",  
                                     RMSE = adj_model_2_RMSE))
RMSE_results %>% knitr::kable()
```

#### Part 4.3 Insights
RMSE has improved to **`r adj_model_2_RMSE`**

### Part 4.4 - Lambda (Regularization rate)
* The lambda (regularization parameter) reduces overfitting of the model
* It reduces the variance of the model's estimated regression parameters
* However, it adds bias to the estimate
* Therefore, try out various lambda values and select the lambda value that produces the smallest RMSE value
```{r, warning=FALSE}
lambdas <- seq(0, 10, 0.25)

rmses <- sapply(lambdas, function(l){
  
  mu <- mean(edx$rating)
  
  b_i <- edx %>% 
    group_by(movieId) %>%
    summarize(b_i = sum(rating - mu)/(n()+l))
  
  b_u <- edx %>% 
    left_join(b_i, by="movieId") %>%
    group_by(userId) %>%
    summarize(b_u = sum(rating - b_i - mu)/(n()+l))
  
  predicted_ratings <- 
    validation %>% 
    left_join(b_i, by = "movieId") %>%
    left_join(b_u, by = "userId") %>%
    mutate(pred = mu + b_i + b_u) %>%
    pull(pred)
  
  return(RMSE(predicted_ratings, validation$rating))
})
```

Add Regularized Model RMSE results to a table for later comparison
```{r, warning=FALSE, echo=FALSE}
RMSE_results <- bind_rows(RMSE_results,
                          data_frame(Model="Regularized Model",  
                                     RMSE = min(rmses)))
```

#### Part 4.4 Insights
RMSE has improved to **`r min(rmses)`**

## STEP 5: EVALUATION
The RMSE values produced by the different models are reproduced in the table below 
```{r, warning=FALSE, echo=FALSE}
RMSE_results %>% knitr::kable()
```

# RESULTS

## Modeling Results & Model Performance
To recap, the goal of this project is to create a machine learning algorithm that predicts movie ratings in the validation dataset as if they were unknown.  

RMSE value is used to evaluate the effectiveness of the models. A lower RMSE value indicates a better fit.

In summary, the lowest RMSE value is **`r min(rmses)`** , achieved from the **Regularized Model**.

# CONCLUSION

## Brief Summary
This report explained the steps in creating a machine learning algorithm to predict movie ratings in the validation dataset from the MovieLens dataset.

The various machine learning algorithms trained on the training dataset was then used to predict the movie ratings on the validation dataset. Ultimately, iterating subsequent models helped to improve the RMSE result.

## Limitations and Future Work
For the purpose of this project, the MovieLens dataset used was a subset of the full dataset.  
To improve results of predicted ratings in future work, one can utilize the full MovieLens dataset, which includes variables like *tags* that have been assigned by users. This informative variable is likely to improve model results.

## Github
This is the link to the GitHub repository containing the reports in PDF format, Rmd format and R script:  
https://github.com/Nicole-Yong/HarvardX-MovieLens